---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# REMLSimulationPaper

<!-- badges: start -->
<!-- badges: end -->

The goal of REMLSimulationPaper is to ...

## Installation

You can install the development version of REMLSimulationPaper from [GitHub](https://github.com/) with:

```{r eval=FALSE, include=TRUE}
# install.packages("devtools")

# install package required for simulation and analysis
devtools::install_github("damian-t-p/halfsibdesign")

# install REMLSimulationPaper
devtools::install_github("damian-t-p/REMLSimulationPaper")
```

## Package structure

All simulation results can be found in `.csv` or `.RData` format in the `/data/` directory.
The figures used in the paper are available in the `/figs/` directory.

To reproduce the simulations and figures, re-build the vignettes by running
```{r eval = FALSE, include=TRUE}
devtools::build_vignettes(pkg = "REMLSimulationPaper")
```

