---
title: "Simulations for Section 3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{compare_REML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

```{r setup}
library(REMLSimulationPaper)
```

```{r}
set.seed(1234)

pacman::p_load(
  tidyverse,
  MASS,
  nlme,
  lme4
)

pacman::p_load_gh(
  "dpavlyshyn/halfsibdesign" # Simulate and fit balanced half-sib designs
)
```

This document contains code for producing simulations and figures used in Section 3.

# Table 1: Comparisons of estimation methods

As described in eq. (3), we set $p = 4, I = 100, J = 3, K = 5$.
```{r}
p <- 4 # number of traits

I <- 100 # number of sires
J <- 3 # number of dams
K <- 5 # number of individuals per line
```

We also take the among-sire, among-dam and among-individual covariance matrices as specified in eq. (4).
That is,
\[
  \Sigma_A = \sigma_A^2 \begin{pmatrix}1&&&\\&1&&\\&&0&\\&&&0\end{pmatrix}, 
  \qquad \Sigma_B = \sigma_B^2 \begin{pmatrix}1&&&\\&0&&\\&&0&\\&&&1\end{pmatrix},
  \qquad \Sigma_E = \sigma_E^2 I,
\]
where we choose $\sigma_A = 5, \sigma_B = 3, \sigma_E = 1$ and $\mu = (1, 2, 3, 4)$.

```{r}
mu <- 1:p

sigma_a <- 5
sigma_b <- 3
sigma_e <- 1

A <- sigma_a^2 * diag(c(1, 1, 0, 0))
B <- sigma_b^2 * diag(c(1, 0, 0, 1))
E <- sigma_e^2 * diag(p)
```

The function `rhalfsib()` from the `dpavlyshyn\halfsibdesign` produces a random realization from the half-sib design with specified intercept, covariance matrices and counts.

We sample one such design and store it in a long data frame.
```{r}
df <- rhalfsib(mu, A, I, B, J, E, K)
head(df)
```

We then compute the stepwise REML and pseudo REML and MANOVA estimates using the corresponding functions contained in `halfsibdesign`.
```{r}
stepreml_fit <- stepreml_2way(df)
pseudoreml_fit <- preml_2way(df)
manova_fit <- manova_2way(df)
```

## Comparison to `lme()`

Next, we fit a mixed-effects model to the data using `lme()` to optimize REML.
The specification for the two-way MANOVA in `lme()` is complicated by the fact that, by default, this function includes uncorrolated homoskedastic error terms.

To explain the specification, we write the model as
\[
y_{ijk,q} = \overbrace{\mu_q}^{\text{fixed effect}} + \overbrace{\alpha_{i,q} + \beta_{ij,q}}^{\text{random effect}} + \overbrace{\epsilon_{ijk,q}}^{\text{residual error}}.
\]

The fixed-effectd are modelled by $y_{ijk,q} \sim \mu_q$ (`fixed = value ~ -1 + trait`).
For the random effects, we estimate the covariances between trait-entries of $\alpha_i, \beta_{ij}$.
That is, we group the random effects at both the sires and the dams level (`random = ~ -1 + trait | sire/dam`).

Last, the variance of each $\epsilon_{ijk,q}$ depends on $q$ only (`weights = varIdent(form = ~ 1 | trait)`), whereas the $\epsilon_{ijk,q}$ are uncorrelated between individuals, and we make no assumptions on the correlations among traits (`correlation = corSymm(form =  ~ 1 | sire/dam/individual)`).
```{r}
lme1 <- lme(
  fixed       = value ~ -1 + trait,
  data        = df,
  random      = ~ -1 + trait | sire/dam,
  weights     = varIdent(form = ~ 1 | trait), # residual variance structure
  correlation = corSymm(form =  ~ 1 | sire/dam/individual), # residual correlation structure
  method      = "REML",
  control     = list(returnObject = TRUE) # don't error if no convergence
  )
```

The `get_covs()` function contained in this package then computes the estimates $\hat{\Sigma}_A, \hat{\Sigma}_B, \hat{\Sigma}_E$ from the output of an `lme` fit.
```{r}
lme_fit <- get_covs(lme1)
```

We then summarize the eigenvalues and REML criteria of the different fits, saving the results to the `/data/` directory.
```{r}
cov_names = c(S1 = "Sigma_E", S2 = "Sigma_B", S3 = "Sigma_A")

df_comparison_eigs <- bind_rows(
  eigen_table(manova_fit) %>% mutate(method = "MANOVA"),
  eigen_table(lme_fit) %>% mutate(method = "REML-LME"),
  eigen_table(stepreml_fit) %>% mutate(method = "Stepwise REML"),
  eigen_table(pseudoreml_fit) %>% mutate(method = "Pseudo REML")
) %>%
  mutate(component = cov_names[component]) %>%
  relocate(component, method) %>%
  arrange(component, method)

write_csv(df_comparison_eigs, "../data/algorithm_comparison_eigenvalues.csv")

df_comparison_eigs
```

```{r}
df_comparison_crit <- bind_rows(
  tibble(method = "MANOVA", reml_crit = manova_fit$reml_crit),
  tibble(method = "REML-LME", reml_crit = lme_fit$reml_crit),
  tibble(method = "Stepwise REML", reml_crit = stepreml_fit$reml_crit),
  tibble(method = "Pseudo REML", reml_crit = pseudoreml_fit$reml_crit)
)

write_csv(df_comparison_crit, "../data/algorithm_comparison_REML_criterion.csv")

df_comparison_crit
```
