---
title: "repeated_samples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{repeated_samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 5
)

# should all simulations be performed? If this is set to FALSE, the simulated 
# data will instead be loaded from the local ./data/ directory.
perform_simulations = FALSE

# should the computed data be stored in a .RData file in ./data/?
save_data = FALSE & perform_simulations
```

```{r setup}
library(REMLSimulationPaper)
```
```{r}
set.seed(1234)

pacman::p_load(
  tidyverse,
  MASS
)

pacman::p_load_gh(
  "dpavlyshyn/halfsibdesign" # Simulate and fit balanced half-sib designs
)

# Set a global theme for all figures in the document
theme_set(theme_bw())
```

# Feasibility of large $q$ for stepwise REML

An important advantage of stepwise REML is that it is can easily be run for relatively large numbers of traits.
In this section, I use $q = 50$, but $q$ can be in the thousands before the pseudo-REML estimate takes more than a second to compute.

We set up the parameters on our now 50-dimensional model:
```{r}
q <- 50 # number of traits

I <- 100 # number of sires
J <- 3 # number of dams
K <- 5 # number of individuals per line

mu <- rep(2, q) # Just a vector of 2s - not really important for the model

sigma_a <- abs(rnorm(q))
sigma_b <- abs(rnorm(q))
sigma_e <- 1

A <- sigma_a^2 * diag(q)
B <- sigma_b^2 * diag(q)
E <- sigma_e^2 * diag(q)
```

As before, we generate our half-sib data and fit a pseudo-REML.
To demonstrate some more complicated begavior, we set $\Sigma_A, \Sigma_B$ to be diagonal matrices with reandom entries.
We also set $q = 50$ to see that the software has no issues with this dimensionality.
```{r}
df <- rhalfsib(mu, A, I, B, J, E, K)
# write_csv(df, "data/halfsib-highdim.csv")

A_reml_est <- stepreml_2way_df(df)

eigs_reml_est <- eigen(A_reml_est$S3, symmetric = TRUE)$values
min(eigs_reml_est)
```

We then plot the eigenvalues of $\hat{\Sigma}_A$ (in red) next to the true eigenvalues of $\Sigma_A$ (in blue):
```{r first-REML}
estimate_names = expression(
  REML = "REML estimate",
  True = "True"~Sigma[A]
)

eig_df <- tibble(
  REML = eigs_reml_est,
  True = sigma_a
) %>%
  pivot_longer(cols = c(REML, True), names_to = "Type", values_to = "Eigenvalue") %>%
  mutate(Type = factor(Type, levels = c("True", "REML")))

ggplot(eig_df, mapping = aes(x = Eigenvalue, fill = Type)) + 
  geom_histogram() + 
  facet_wrap(
    ~Type,
    labeller = label_bquote(.(estimate_names[as.character(Type)]))) + 
  labs(
    title = expression("Histograms of eigenvalues of"~Sigma[A]~"and its REML estimate"),
    y = "Count"
  ) +
  theme(legend.position = "none")

ggsave("../figs/first-REML.pdf", device = "pdf")
```

## Pairwise REML

Last, we investigate the strategy of computing pairwise REML estimates for each pair of traits.
We create a new estimate $\hat{\Sigma}_{A,2}$ such that the diagonal entries $[\hat{\Sigma}_{A,2}]_{ii}$ are computed by doing a single-trait analysis on trait $i$ and the off-diagonal entries $[\hat{\Sigma}_{A,2}]_{ij}$ are computed by doing a 2-trait analysis on the traits $(i,j)$.

I use the pseudo-REML estimates for the 2-trait analyses, since we need to fill in 1250 entries of a covariance matrix, and even in the 2-dimensional case, fitting a generic mixed model is too slow for this.
However, we can still compare $\hat{\Sigma}_{A,2}$ with the REML esimate $\hat{\Sigma}_{A,1}$ to see how the eigenvalues produced by the paiwiise procedure differ from those done by computing the whole stepwise REML estimate at once.

```{r eval = perform_simulations}
A_pairwise_est <- matrix(nrow = q, ncol = q)

for(r in 1:q) {
  # fill in diagonal
  df_sub <- filter(df, trait == r)
  A_sub_est <- stepreml_2way_df(df_sub)$S3
  
  A_pairwise_est[r, r] <- A_sub_est[1,1]
  
  if(r < q) {
    for(s in (r+1):q) {
      df_sub <- filter(df, trait %in% c(r, s))
      A_sub_est <- stepreml_2way_df(df_sub)$S3
      
      A_pairwise_est[r, s] <- A_sub_est[1,2]
      A_pairwise_est[s, r] <- A_pairwise_est[r, s]
    }
  }
}
```

```{r eval = !perform_simulations}
A_pairwise_est <- read.table("../data/pairwise-covariances.csv", sep = ",") %>%
  unname() %>%
  as.matrix()
```

```{r eval = save_data}
write.table(
  A_pairwise_est,
  "../data/pairwise-covariances.csv",
  sep = ",",
  row.names = FALSE,
  col.names = FALSE)
```

Plotting these eigenvalues, we see that the eigenvalues of $A_2$ are qualitatively similar to those of $A_1$.

```{r}
eigs_pairwise_est <- eigen(A_pairwise_est, symmetric = TRUE)$values
```

To make this comparison clearer, let $\lambda_1 > \dotsc > \lambda_q$ the the eigenvalues of $A_1$ and $\mu_1 > \dotsc > \mu_q$ be the eigenvalues of $A_2$.

In the following figure, we plot $\mu_i - \lambda_i$ against $\lambda_i$.
```{r eval-diffs}
ggplot(mapping = aes(x = eigs_reml_est, y = eigs_pairwise_est - eigs_reml_est)) + 
  geom_point() + 
  labs(
    title = expression("Comparison of eigenvalues of"~Sigma["A,R"]~"and"~Sigma["A,P"]),
    x = expression("Eigenvalues"~lambda["i, R"]~"of"~Sigma["A,R"]),
    y = expression(lambda["i, R"] - lambda["i, P"])
  )

ggsave("../figs/eval-diffs.pdf", device = "pdf")
```


# Repeated sampling

With the improved algorithm for computing REML esimates of $\Sigma_A, \Sigma_B, \Sigma_E$ in the balanced case, investigation of large eigenvalues of these matrices throughrepeated sampling becomes much more tractable.

```{r}
repeated_sample <- function(q_vals, I, J, K, n_reps, methods, make_sigmas, make_mu, max_idx=5) {
  # output data frame
  res_df <- tibble(
    q = integer(),
    rep = integer(),
    level = character(),
    eig_idx = integer(),
    method = character(),
    pop_eig = numeric(),
    sample_eig = numeric()
  )
  
  for(q in q_vals) {
    A <- make_sigmas$A(q)
    B <- make_sigmas$B(q)
    E <- make_sigmas$E(q)
    
    mu_vec <- make_mu(q)
    
    for(rep in 1:n_reps) {
      
      df <- rhalfsib(mu_vec, A, I, B, J, E, K)
      
      for(method in names(methods)) {
        reml_est <- methods[[method]](df)
        #reml_est <- manova_2way_df(df)
        
        A_eigs <- eigen(reml_est$S3, symmetric = TRUE, only.values = TRUE)$values
        
        df_A <- tibble(
          q = q,
          rep = rep,
          level = "A",
          eig_idx = 1:max_idx,
          method = method,
          pop_eig = sigma_a^2,
          sample_eig = A_eigs[1:max_idx]
        )
        
        B_eigs <- eigen(reml_est$S2, symmetric = TRUE, only.values = TRUE)$values
        
        df_B <- tibble(
          q = q,
          rep = rep,
          level = "B",
          eig_idx = 1:max_idx,
          method = method,
          pop_eig = sigma_b^2,
          sample_eig = B_eigs[1:max_idx]
        )
        
        E_eigs <- eigen(reml_est$S1, symmetric = TRUE, only.values = TRUE)$values
        
        df_E <- tibble(
          q = q,
          rep = rep,
          level = "E",
          eig_idx = 1:max_idx,
          method = method,
          pop_eig = sigma_e^2,
          sample_eig = E_eigs[1:max_idx]
        )
        
        res_df <- bind_rows(res_df, df_A, df_B, df_E)
      }
      
    }
  }
  
  return(res_df)
}
```

## Distribution of large eigenvalues

First, we investigate the largest eigenvalues of REML estimates with increasing numbers of traits.

In this example, we simulate half-sibling designs with  $\Sigma_A = I, \Sigma_B = 2I, \Sigma_E = I, \mu = (2, \dotsc, 2)$ with $I = 100, J= 3, K = 5$.
We perform 50 such simulations for each of $q = 10, 20, \dotsc, 100$ and plot the results.

```{r eval = perform_simulations}
# experiment parameters
q_vals <- seq(10, 100, by=10)
n_reps <- 50
max_idx <- 5

methods <- list(
  #MANOVA = manova_2way_df,
  REML = stepreml_2way_df
)

# model parameters
sigma_a <- 1
sigma_b <- 2
sigma_e <- 1

I <- 100 
J <- 3 
K <- 5 

mu <- 2

make_sigmas <- list(
  A = function(q) {sigma_a^2 * diag(q)},
  B = function(q) {sigma_b^2 * diag(q)},
  E = function(q) {sigma_e^2 * diag(q)}
)

make_mu <- function(q) {rep(mu, q)}

res_df <- repeated_sample(q_vals, I, J, K, n_reps, methods, make_sigmas, make_mu, max_idx=5)
```

```{r eval = !perform_simulations}
res_df <- read_csv(
  "../data/REML-p50-eigenvalues.csv",
  col_types = c("i", "i", "c", "i", "c", "d", "d")
  )
```

```{r eval = save_data}
write_csv(res_df, "../data/REML-p50-eigenvalues.csv")
```

In the following plot, we show the means of the top 5 eigenvalues of the REML estimates for each of $\Sigma_A, \Sigma_B,\Sigma_E$.
Additionally, the corresponding population eigenvalues (1, 2, 1 respectively) are shown in a thick red line.

We see that as $q$ increases, the largest eigenvalues become increasingly biased.
```{r index}
cov_names = expression(
  A = Sigma[A],
  B = Sigma[B],
  E = Sigma[E]
)

summ_df <- res_df %>%
  group_by(q, level, method, eig_idx) %>%
  summarize(
    eig_mean = mean(sample_eig),
    eig_sd = sd(sample_eig),
    pop_eig = mean(pop_eig)
  )

summ_df %>%
  ggplot(aes(x = q, y = eig_mean, color = ordered(eig_idx))) + 
  facet_wrap(
    ~ level,
    scales = "free_y",
    labeller = label_bquote(.(cov_names[level]))) + 
  geom_line() + 
  geom_hline(aes(yintercept = pop_eig), color = "red", size = 1.5) + 
  geom_point() + 
  labs(
    title    = "Largest eigenvalues of REML fit",
    subtitle = "Mean of 50 replicates of top-5 eigenvalues",
    x        = "Number of traits (p)",
    y        = "REML eigenvalue",
    color    = "Index (i)"
  )

ggsave("../figs/index.pdf", device = "pdf")
```

## Difference between REML and MANOVA estimates

Another question of interest is how close the large eigenvalues of the REML estimates are to the corresponding MANOVA eigenvalues.
In the previous example, the eigenvalues of $\Sigma_A, \Sigma_B, \Sigma_E$ were large enough that the REML estimates coincided with the MANOVA estimates.

To see a more pathological example, we repeat the simulations of the previous sub-section, but now with half of the diagonals of $\Sigma_A$ and $\Sigma_B$ set to zero at random.
Now, the smallest eigenvalues of the MANOVA estimates of $\Sigma_A$ and $\Sigma_B$ are negative and so must me corrected.

To examine the discrepancies between the two estimates, we plot the mean and standard deviation of the differnces $\lambda_i(\hat{\Sigma}_{k, \text{REML}}) - \lambda_i(\hat{\Sigma}_{k, \text{MANOVA}})$ for $i = 1, \dotsc, 5, k = A, B, E$.

```{r eval = perform_simulations}
# experiment parameters
q_vals <- seq(10, 70, by=10)
n_reps <- 50
max_idx <- 5

methods <- list(
  MANOVA = manova_2way_df,
  REML = stepreml_2way_df
)

# model parameters
sigma_a <- 1
sigma_b <- 2
sigma_e <- 1

I <- 100 
J <- 3 
K <- 5 

mu <- 2


make_sigmas <- list(
  A = function(q) {sigma_a^2 * diag(rbinom(q, 1, 0.5))},
  B = function(q) {sigma_b^2 * diag(rbinom(q, 1, 0.5))},
  E = function(q) {sigma_e^2 * diag(q)}
)

make_mu <- function(q) {rep(mu, q)}

comp_df <- repeated_sample(q_vals, I, J, K, n_reps, methods, make_sigmas, make_mu, max_idx=5)
```
```{r eval = !perform_simulations}
comp_df <- read_csv(
  "../data/REML-MANOVA-eigenvalue-comparison.csv",
  col_types = c("i", "i", "c", "i", "c", "d", "d")
)
```

```{r eval = save_data}
write_csv(comp_df, "../data/REML-MANOVA-eigenvalue-comparison.csv")
```

We see that the largest eigenvalues of  $\hat{\Sigma}_A$ are adjusted up slightly, whereas the largest eigenvalues of $\hat{\Sigma}_B$ are adjusted down substantially.

```{r}
comp_summary <- comp_df %>%
  pivot_wider(names_from = "method", values_from = "sample_eig") %>%
  mutate(est_diff = REML - MANOVA) %>%
  group_by(q, level, eig_idx) %>%
  summarize(
    diff_mean = mean(est_diff),
    diff_sd = sd(est_diff)
  )
```
```{r e-diff}
comp_summary %>%
  ggplot(aes(x = q, y = diff_mean, color = ordered(eig_idx))) +
  geom_line() +
  geom_point() + 
  facet_wrap(
    ~ level,
    labeller = label_bquote(.(cov_names[level]))) +
  labs(
    title = "Difference between REML and MANOVA eigenvalues",
    subtitle = "Mean difference of 50 replicates of top-5 eigenvalues",
    y = expression(lambda["i,R"] - lambda["i,M"]),
    x = "Number of traits (p)",
    color = "Index (i)"
  )

ggsave("../figs/e-diff.pdf", device = "pdf")
```
